---
title: "Medicaid Data Trends"
author: "Patrick Carr"
date: "02-07-2023"
output: 
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE, include = FALSE}
install.packages('pacman', repos = "http://cran.us.r-project.org")
library(pacman)
p_load(tidyverse, janitor, lubridate, dplyr, ggplot, tinytex)

# Import and format data on Medicaid eligibility cutoffs
eligibility <- read_csv("Raw Data/eligibility.csv", 
                        col_names = c("Year", "Percent"), skip = 2)
eligibility$Year <- as.Date(as.character(eligibility$Year), format = "%Y")

```

## Data Exploration
This exploration will look at the impact of the increase in Medicaid eligibility as a percentage of the federal poverty line.

I will be exploring trends in four separate data sets. First, I will look at the number of emergency room visits per thousand Americans, pear year, as measured by the [Kaiser Family Foundation](https://www.kff.org/other/state-indicator/emergency-room-visits-by-ownership/?activeTab=graph&currentTimeframe=0&startTimeframe=16&selectedDistributions=total&selectedRows=%7B%22wrapups%22:%7B%22united-states%22:%7B%7D%7D%7D&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D):
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ervisits <- read_csv("Raw Data/er-visits.csv", 
                     col_names = c("Year", "Visits"), skip = 2)
ervisits$Year <- as.Date(as.character(ervisits$Year), format = "%Y")

print("Emergency Room Visits Per Thousand Americans")
summary(ervisits$Visits)
```


Second, I will explore the percentage of Americans who forego medical care due to cost per year with data from the [State Health Data Compare Assistance Center](https://statehealthcompare.shadac.org/landing/178/percent-of-adults-who-could-not-get-medical-care-when-needed-due-to-cost-by-total-2011-to-2021) :
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Import and format data on Americans foregoing medical care due to cost
forego <- read_csv("Raw Data/foregoing-care.csv", 
                   col_names = c("Abv", "Location", "Year", "Format", "Percent", "MOE"), skip = 6)
forego <- subset(forego, select = c("Year", "Percent"))
forego$Year <- as.Date(as.character(forego$Year), format = "%Y")

print("Percentage of Americans Reporting Foregoing Medical Care Due to Cost")
summary(forego$Percent)
```

Third, I will explore the net assets and debts of Americans in the lowest 20th percentile of income earners per year using data from the [Federal Reserve's Consumer Finances Survey](https://www.federalreserve.gov/econres/scf/dataviz/scf/chart/#series:Before_Tax_Income;demographic:inccat;population:1;units:median;range:1989,2019):
```{r, echo = FALSE, message = FALSE, warning = FALSE}
finances <- read_csv("Raw Data/finances.csv", 
                     col_names = c("Year", "Status", "Income", "Assets", "Debt"), skip = 5)
finances <- subset(finances, select = c("Year", "Assets", "Debt"))
finances$Year <- as.Date(as.character(finances$Year), format = "%Y")

print("Assets and Debts of Lowest 20% of American Earners")
summary(finances[,(2-3)])
```

Finally, I will explore the health statuses of Americans per year using data from the [Center for Disease Control](https://www.cdc.gov/nchs/hus/topics/health-status.htm#explore-data). This data reports the percentage of Americans reporting "fair" or "poor" health by year.

This dataset separates health statuses by demographics, including gender:
```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Import, truncate, and format data on health status survey data (lower value = better health)
hstatus <- read_csv("Raw Data/health-status.csv",
                    col_names = c("Year", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", 
                                  "2013", "2014", "2015", "2016", "2017", "2018", "2019"), skip = 5)
hstatus <- as.data.frame(t(hstatus[1:22, 1:15]))
hstatus <- hstatus |>  row_to_names(row_number = 1)
hstatus <- cbind(Year = rownames(hstatus), hstatus)
rownames(hstatus) <- 1:nrow(hstatus)
hstatus$Year <- as.Date(as.character(hstatus$Year), format = "%Y")
hstatus[,2:22] = apply(hstatus[,2:22], 2, function(x) as.numeric(as.character(x)))
colnames(hstatus) <- c("Year", "All_Ages_adj", "All_Ages_crude", "Under18", "Under6", "6to17", "18to44",
                       "18to24", "25to44", "45to54", "55to64", "Over65", "65to74", "Over75", "Male", 
                       "Female", "White", "Black", "Hispanic", "Under100pct", "100to199pct", "200to399pct", 
                       "Over400pct")

print("Health Status by Gender")
summary(hstatus[c("Male", "Female")])
```

Race:
```{r, echo = FALSE, message = FALSE, warning = FALSE}
print("Health Status by Race")
summary(hstatus[c("White", "Black", "Hispanic")])
```

And income level. These variables report health statuses for Americans based on their income as a percentage of the federal poverty line (e.g., the variable "100to199pct" is the health status of individuals earning between 100% and 199% of the poverty line).
```{r, echo = FALSE, message = FALSE, warning = FALSE}
print("Health Status by Income Level")
summary(hstatus[c("Under100pct", "100to199pct", "200to399pct", "Over400pct")])
```





## Data Trends

This trend shows the average percentage of the poverty line to qualify for Medicaid coverage (this data was retrieved from the [Kaiser Family Foundation](https://www.kff.org/medicaid/state-indicator/medicaid-income-eligibility-limits-for-parents/?currentTimeframe=0&selectedDistributions=january-2002--april-2003--july-2004--july-2005--july-2006--january-2008--january-2009--december-2009--january-2011--january-2012--january-2013--january-2014--january-2015--january-2016--january-2017--january-2018--january-2019--january-2020--january-2021--january-2022&selectedRows=%7B%22wrapups%22:%7B%22united-states%22:%7B%7D%7D%7D&sortModel=%7B%22colId%22:%22Location%22,%22sort%22:%22asc%22%7D)). In 2013, the eligibility leaps up to 138% due to the implementation of the Affordable Care Act.

```{r , echo = FALSE, message = FALSE, warning = FALSE}
eligibility |> 
  ggplot() +
  geom_point(aes(x = Year, y = Percent), size = 2.5, color = "#1d65a0") +
  ylim(0, 1.4) +
  xlab("Year") + 
  ylab("Eligibility as % of Poverty Line") +
  ggtitle("Medicaid Eligibility Trend") +
  theme(panel.background = element_rect(fill = '#fcf9f7', color = 'black'),
        panel.grid.major = element_line(color = '#818a92', linetype = 'dotted'),
        panel.grid.minor = element_line(color = '#818a92', linetype = 'dotted'),
        axis.text.x = element_text(color = 'black', size = 11),
        axis.text.y = element_text(color = 'black', size = 11),
        axis.title.x = element_text(color = 'black', size = 12.5),
        axis.title.y = element_text(color = 'black', size = 12.5),
        plot.title = element_text(size = 14, face = "bold"))
```

The following data trends will attempt to identify trends that may have been driven by this leap upwards to 138%.

First, let's look at trend in hospital visits:
```{r, echo = FALSE, message = FALSE, warning = FALSE}
ervisits |> ggplot() +
  geom_line(aes(x = Year, y = Visits), size = 1.5, color = '#7d0540') +
  geom_vline(aes(xintercept = as.Date(Year[9])), linetype = 'twodash', size = 1.5, color = "#1d65a0") +
  ylim(200, 450) +
  xlab("Year") + 
  ylab("Number of ER Visits") +
  ggtitle("Emergency Room Visits per Thousand Americans") +
  theme(panel.background = element_rect(fill = '#fcf9f7', color = 'black'),
        panel.grid.major = element_line(color = '#818a92', linetype = 'dotted'),
        panel.grid.minor = element_line(color = '#818a92', linetype = 'dotted'),
        axis.text.x = element_text(color = 'black', size = 11),
        axis.text.y = element_text(color = 'black', size = 11),
        axis.title.x = element_text(color = 'black', size = 12.5),
        axis.title.y = element_text(color = 'black', size = 12.5),
        plot.title = element_text(size = 14, face = "bold"))

```

Then let's look at Americans foregoing medical care:
```{r, echo = FALSE, message = FALSE, warning = FALSE}
forego |> ggplot() +
  geom_line(aes(x = Year, y = Percent), size = 1.5, color = "#f19633") +
  geom_vline(aes(xintercept = as.Date(Year[9])), linetype = 'twodash', size = 1.5, color = "#1d65a0") +
  ylim(0, 0.17) +
  xlab("Year") +
  ylab("% Foregoing") +
  ggtitle("Percentage Foregoing Medical Care Due to Cost") +
  theme(panel.background = element_rect(fill = '#fcf9f7', color = 'black'),
        panel.grid.major = element_line(color = '#818a92', linetype = 'dotted'),
        panel.grid.minor = element_line(color = '#818a92', linetype = 'dotted'),
        axis.text.x = element_text(color = 'black', size = 11),
        axis.text.y = element_text(color = 'black', size = 11),
        axis.title.x = element_text(color = 'black', size = 12.5),
        axis.title.y = element_text(color = 'black', size = 12.5),
        plot.title = element_text(size = 14, face = "bold"))
```

Next we'll look at assets and debts of the poorest Americans:
```{r, echo = FALSE, message = FALSE, warning = FALSE}
finances |> ggplot(aes(x = Year)) +
  geom_point(aes(y = Debt, color = '#fc5c88'), size = 2.75) +
  geom_point(aes(y = Assets, color = "black"), size = 2.75) +
  geom_vline(aes(xintercept = as.Date(Year[5])), linetype = 'twodash', size = 1.5, color = "#1d65a0") +
  xlab("Year") + 
  ylab("Amount in Thousands") +
  ggtitle("Debts and Assets of Lowest 20% of American Earners") +
  scale_color_identity(name = '',
                       breaks = c('black', '#fc5c88'),
                       labels = c("Assets", "Debt"),
                       guide = 'legend') +
  theme(panel.background = element_rect(fill = '#fcf9f7', color = 'black'),
        panel.grid.major = element_line(color = '#818a92', linetype = 'dotted'),
        panel.grid.minor = element_line(color = '#818a92', linetype = 'dotted'),
        axis.text.x = element_text(color = 'black', size = 11),
        axis.text.y = element_text(color = 'black', size = 11),
        axis.title.x = element_text(color = 'black', size = 12.5),
        axis.title.y = element_text(color = 'black', size = 12.5),
        plot.title = element_text(size = 14, face = "bold"))
```

Health status of those under the poverty line:
```{r, echo = FALSE, message = FALSE, warning = FALSE}
hstatus |> ggplot(aes(x = Year)) + 
  geom_line(aes(y = Under100pct), color = '#00853f', size = 1.5) +
  geom_vline(aes(xintercept = as.Date(Year[9])), color = "#1d65a0", linetype = 'twodash', size = 1.5) +
  ylim(18,22) +
  xlab("Year") + 
  ylab("Percentage") +
  ggtitle("Percentage of Impoverished Reporting Fair or Poor Health") + 
  theme(legend.position="none") +
  theme(panel.background = element_rect(fill = '#fcf9f7', color = 'black'),
        panel.grid.major = element_line(color = '#818a92', linetype = 'dotted'),
        panel.grid.minor = element_line(color = '#818a92', linetype = 'dotted'),
        axis.text.x = element_text(color = 'black', size = 11),
        axis.text.y = element_text(color = 'black', size = 11),
        axis.title.x = element_text(color = 'black', size = 12.5),
        axis.title.y = element_text(color = 'black', size = 12.5),
        plot.title = element_text(size = 14, face = "bold"))
```

