---
title: "Medicaid Data Exploration"
author: "Patrick Carr"
date: "02-07-2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Exploration
This exploration will look at the impact of the increase in Medicaid eligibility as a percentage of the federal poverty line.

I will be exploring trends in five separate data sets. First, I will look at the number of emergency room visits per thousand Americans:

```{r}
summary(ervisits$Visits)
```

## Data Trends

This trend shows the average percentage of the poverty line to qualify for Medicaid coverage. In 2013, the data leaps up to 138% due to the implementation of the Affordable Care Act.

```{r pressure, echo=FALSE}
eligibility |> 
  ggplot() +
  geom_point(aes(x = Year, y = Percent), size = 2.5, color = "#1d65a0") +
  ylim(0, 1.4) +
  xlab("Year") + 
  ylab("Eligibility as % of Poverty Line") +
  ggtitle("Medicaid Eligibility Trend") +
  theme(panel.background = element_rect(fill = '#fcf9f7', color = 'black'),
        panel.grid.major = element_line(color = '#818a92', linetype = 'dotted'),
        panel.grid.minor = element_line(color = '#818a92', linetype = 'dotted'),
        axis.text.x = element_text(color = 'black', size = 11),
        axis.text.y = element_text(color = 'black', size = 11),
        axis.title.x = element_text(color = 'black', size = 12.5),
        axis.title.y = element_text(color = 'black', size = 12.5),
        plot.title = element_text(size = 14, face = "bold"))
```

